---
title: "Exploratory Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{exploratory-data-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Get the training data, see `titanic_train` for more info.

```{r, message=FALSE}
library(iceberg)
library(data.table)
library(ggplot2)

train <- iceberg::titanic_train
train <- as.data.table(train)
str(train)
```


```{r}
#for plots
train$PassengerStatus <- as.factor(ifelse(train$Survived == 1, 'Survived','Died'))
```
## Passanger Analysis

### Who Survived?

The story goes, that women and children were saved first.

```{r}
ggplot(train, aes(x=Sex, y=Age, fill=PassengerStatus)) + geom_bar(position = 'fill', stat='identity') +
    labs(title = "Proportion of Survival by Gender and Age") +
    scale_fill_brewer(palette="Dark2")
```

It does appear that a lot more women and younger males were saved.

```{r}
ggplot(train, aes(x=Sex, y=Age, fill=PassengerStatus)) + geom_bar(position = 'fill', stat='identity') +
    facet_wrap(.~Pclass) +
    labs(title = "Proportion of Survival by Gender and Age",
         subtitle = "Faceted by Passenger Class") +
    scale_fill_brewer(palette="Dark2")
```
Dang, you don't want to be in 3rd class, survival rate for male and female is less than 50%.

```{r}
ggplot(train, aes(x=Sex, y=Age, fill=PassengerStatus)) + geom_bar(position = 'fill', stat='identity') +
    facet_wrap(.~Parch) +
    labs(title = "Proportion of Survival by Gender and Age",
         subtitle = "Faceted by Passenger Parch") +
    scale_fill_brewer(palette="Dark2")
```

```{r}
ggplot(train, aes(x=Sex, y=Age, fill=PassengerStatus)) + geom_bar(position = 'fill', stat='identity') +
    facet_wrap(.~Embarked) +
    labs(title = "Proportion of Survival by Gender and Age",
         subtitle = "Faceted by Passenger Embarked") +
    scale_fill_brewer(palette="Dark2")
```

There are some null age values...

```{r}
table(is.na(train$Age))
prop.table(table(is.na(train$Age)))
```

almost 20% of the observations are missing an age value. If I were to impute age based on the average it would look like the chart below.

```{r}
train <- iceberg::impute_passenger_age(train)
train$AgeGroup <- cut(train$Age, breaks = c(0,17,30,40,50,65,80), labels = c('<17','18-30','31-40','41-50','51-65','>=66'))

ap <- ggplot(train, aes(x=Age, fill=Sex)) + geom_density(position = 'stack', alpha=0.5)

ap
```

```{r}
ap + facet_wrap('Embarked')
  
prop.table(table(train$Sex,train$Embarked))
```
