---
title: "Modeling Titanic Survival"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Titanic Survival}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Details of this Kaggle challenge can be found [here](https://www.kaggle.com/c/titanic), the challenge is to accurately predict whether a passenger survived the shipwreck or not. The response variable is binary and since a passenger cannot partially survive the response variable will be either 1 or 0. This document will use the [Stan](https://mc-stan.org/) programming language and logistic regression to attack the challenge.

```{r, message=FALSE}
library(rstan)
library(magrittr)
library(ggplot2)
library(iceberg)
library(bayesplot)

train <- iceberg::titanic_train
test <- iceberg::titanic_test

train_idx <- sample(nrow(train), nrow(train)*0.8)
test_idx <- setdiff(seq_len(nrow(train)), train_idx)

str(train)
```

It's a common anecdote that women and children were the first passengers saved so the first model will only use age and gender as predictors. Since there is a substantial number of missing age values, nulls will be imputed, see the `Impute Passenger Age` function for the details of the imputation methodology or the `Exploratory Data Analysis` vignette to see the before after.

```{r}
train <- iceberg::impute_passenger_age(train)
test <- iceberg:: impute_passenger_age(test)
```

## Modeling

The `Stan` program fits the model but also accepts a test dataset which will be used to make predictions. For more details on this approach see the `Stan` [manual](https://mc-stan.org/docs/2_24/stan-users-guide/prediction-forecasting-and-backcasting.html).

```{r, message=FALSE}
simple_model <- iceberg::age_gender_stan(
                                          age=train[train_idx,'Age'],
                                          sex=2,
                                          sex_idx=as.integer(train[train_idx,'Sex']),
                                          survived=train[train_idx,'Survived'],
                                          test_age=train[test_idx,'Age'],
                                          test_sex_idx=as.integer(train[test_idx,'Sex']), seed=1234)
```

The model estimates an intercept for each `Sex` separately but shares the coefficient on age. The prior on `beta` is a standard normal distribution centered at 0 with a standard deviation of 1.

```{r}
print(simple_model, pars=c('alpha','beta'))
```


```{r, message=FALSE}
samp <- rstan::extract(simple_model)

mcmc_areas(
  simple_model, 
  pars = c("alpha[1]","alpha[2]"),
  prob = 0.8, # 80% intervals
  prob_outer = 0.99, # 99%
  point_est = "mean",
) +
  labs(title = 'Model Intercepts for Male and Female Passengers')
```

`alpha[1]` is the posterior distribution for the intercept related to female passengers on the ship. It's clear that gender plays an important role in determining survival. The shaded regions correspond to 80% uncertainty intervals.

```{r}
post_pred_df <- as.data.frame(samp$y_new)
bayesplot::ppc_stat(y = train[test_idx,"Survived"], yrep = as.matrix(post_pred_df), stat = mean, binwidth = .01)
bayesplot::ppc_stat_grouped(y = train[test_idx,"Survived"], yrep = as.matrix(post_pred_df), stat = mean, group = train[test_idx,"Sex"], binwidth = 0.03)
```

The plot above is comparing the predictions to the actual mean of the test dataset. The model is close but there's still a some variation the model is missing and individual predictions are still in need of review and for the Kaggle challenge we care about accuracy.

```{r}
mean(apply(samp$y_new, 2, median) %>% round(0) == train[test_idx,'Survived'])
```

Accuracy against a test set is about `r mean(apply(samp$y_new, 2, median) %>% round(0) == train[test_idx,'Survived'])` percent. There seems to be more to the story we need to consider to improve accuracy to a reasonable level.

Predict on the Kaggle challenge test set.

```{r}
predict_model <- iceberg::age_gender_stan(age=train[train_idx,'Age'],
                                          sex=2,
                                          sex_idx=as.integer(train[train_idx,'Sex']),
                                          survived=train[train_idx,'Survived'],
                                          test_age=test$Age,
                                          test_sex_idx=as.integer(test$Sex)
                                         , seed=1234
                                         )
```

Export the predictions and push to Kaggle.

```{r}
pred_samp <- rstan::extract(predict_model)
test$Survived <- apply(pred_samp$y_new, 2, median)
#write.csv(test[,c("PassengerId","Survived")] ,file="inst/extdata/predict_20201002.csv", row.names = FALSE)
```

### Model Iteration

The anecdote we started with was that women and children were saved first and based on the initial model above this appears to be a true statement. To build on the anecdote, passengers with a higher societal standing also were given priority over lower-class passengers. To build on the model above, I'll add passenger class as an additional input into the model and see how much improvement is achieved.

```{r}
class_mod <- iceberg::age_gender_class_stan(age=train[train_idx,'Age'],
                                    pclass=train[train_idx,'Pclass'],
                                    sex=2,
                                    sex_idx=as.integer(train[train_idx,'Sex']),
                                    survived=train[train_idx,'Survived'],
                                    test_age=train[test_idx,'Age'],
                                    test_pclass=train[test_idx,'Pclass'],
                                    test_sex_idx=as.integer(train[test_idx,'Sex'])
                                    , seed=124
                                    )
```

```{r}
class_fit <- rstan::extract(class_mod)
print(class_mod, pars=c("alpha","beta","beta_class"))
```

Checking the accuracy against my test partition.

```{r}
mean(apply(class_fit$y_new, 2, median) %>% round(0) == train[test_idx,'Survived'])
```

The accuracy seems to have barely improved...

```{r, message=FALSE}
bayesplot::mcmc_intervals(class_mod, pars=c("alpha[1]","alpha[2]"), prob=0.8) +
  ggplot2::labs(title = 'MCMC Intervals on Gender Based Intercepts') +
  ggplot2::scale_y_discrete(labels = c("alpha[1]"="Male", "alpha[2]"="Female"))

bayesplot::mcmc_intervals(class_mod, pars=c("beta","beta_class"), prob=0.8) +
  ggplot2::labs(title = 'Coefficients on Age and Passenger Class')
```

```{r}
bayesplot::ppc_stat_grouped(y = train[test_idx,"Survived"], yrep = class_fit$y_new, stat = mean, group = train[test_idx,"Sex"], binwidth = 0.03)
```

```{r}
class_predict <- iceberg::age_gender_class_stan(age=train[train_idx,'Age'],
                                    pclass=train[train_idx,'Pclass'],
                                    sex=2,
                                    sex_idx=as.integer(train[train_idx,'Sex']),
                                    survived=train[train_idx,'Survived'],
                                    test_age=test$Age,
                                    test_pclass=test$Pclass,
                                    test_sex_idx=as.integer(test$Sex)
                                    , seed=124
                                    )
```

Export the predictions and push to Kaggle.

```{r}
class_pred <- rstan::extract(class_predict)
test$Survived <- apply(class_pred$y_new, 2, median)
#write.csv(test[,c("PassengerId","Survived")] ,file="~/projectrepos/iceberg/inst/extdata/predict_age_class.csv", row.names = FALSE)
```

## Stan Models

The first model is the age and gender model that creates separate intercepts for each gender.

```{r engine='bash', comment='', echo=FALSE}
cat ~/projectrepos/iceberg/inst/stan/age_gender_model.stan
```

The second model incorporates the passenger class status, extending the model from above.

```{r engine='bash', comment='', echo=FALSE}
cat ~/projectrepos/iceberg/inst/stan/age_gender_class_model.stan
```

## References

* Stan Development Team. 2018. RStan: the R interface to Stan. R package version 2.17.3.   http://mc-stan.org
* Stan Development Team. 2018. Stan Modeling Language Users Guide and Reference Manual, Version 2.18.0.   http://mc-stan.org
